Parallel Multipoint Approximation Method for Large-Scale Optimization Problems\thanks{This study was supported by the Russian Science Foundation, project No.\,16-11-10150.}

### 2018-05-07 2018/PAVT_Toropov/PAVT2018_MAM.tex
After checking

% first the title is needed
\title{Parallel Multipoint Approximation Method for Large Scale Optimization Problems 
\thanks{This study was supported by the Russian Science Foundation, project No. 16-11-10150.}}

@@

\author{ Victor P. Gergel%(ORCID 0000-0002-4013-2329)
$^1$ \and
Konstantin A. Barkalov%(ORCID 0000-0001-5273-2471)
$^1$ \and
Evgeny A. Kozinov%(ORCID 0000-0001-6776-0096)
$^1$ \and \\
Vassili V. Toropov%(ORCID 0000-0002-7017-5055)
$^{1,2}$
}
%
\authorrunning{V.P. Gergel \and K.A. Barkalov \and E.A.Kozinov \and V.V. Toropov}
% (feature abused for this document to repeat the title also on left hand pages)

@@

\begin{abstract}
The paper presents a new development in the Multipoint Approximation Method (MAM) that makes it capable of handling large scale problems. The approach relies on approximations built in the space of design variables within the iterative trust-region-based framework of MAM. 
With the purpose of solving the problems of high dimensionality in a reasonable time, a parallel variant of the Multipoint Approximation Method (PMAM) has been developed. It was supposed that the values of the objective function and of the constraints were computed using the distributed memory (on several cluster nodes), whereas the optimization module run on a single node using the shared memory. The numerical experiments have been carried out on a benchmark example of structural optimization.

@@

In the present paper, the multipoint approximation method (MAM \cite{Toropov1989,Toropov1992,ToropovFilatov1993}) and its application to the large scale optimization problems are considered. In problems with a large (in the order of hundreds) number of design variables MAM proved to be efficient, e.g. in turbomachinery applications \cite{ShahparPolynkinToropov2008,PolynkinToropovShahpar2008,PolynkinToropovShahpar2010}. This method is an iterative optimization technique based on mid-range approximations built in trust regions. A trust region is a sub-domain of the design space in which a set of design points, produced according to a small-scale design of experiments (DoE), are evaluated. These and a subset of previously evaluated design points are used to build metamodels of the objective and constraint functions that are considered to be valid within a current trust region. The trust region will then translate and change size as optimization progresses. The trust region strategy has gone through several stages of development to account for the presence of numerical noise in the response function values \cite{KeulenToropovMarkine1996,ToropovKeulenMarkine1996} and occasional simulation failures \cite{ToropovMarkineHolden1999}. The mid-range approximations used in the trust regions, as originally suggested in \cite{Toropov1989} for structural optimization problems, are intrinsically linear functions (i.e. nonlinear functions that can be led to a linear form by a simple transformation) for individual sub-structures, and assembly of them for the whole structure. This was enhanced by the use of gradient-assisted metamodels \cite{ToropovFilatov1993}, use of simplified numerical models that is also termed a multi-fidelity approach,\cite{ToropovMarkine1996} and the use of analytical models derived by Genetic Programming \cite{ToropovAlvarez1998}. One of the recent developments \cite{PolynkinToropov2012} involved the use of approximation assemblies, i.e. a two stage approximation building process that is conceptually similar to the original one used in \cite{Toropov1989} but is free from the limitation that lower level approximations are linked to individual substructures.

The Moving Least Squares Method (MLSM) was proposed in \cite{LancasterSalkauskas1981} for smoothing and interpolation of scattered data and later used in the mesh-free form of the finite element method (FEM) \cite{Liszka1984}. As suggested in \cite{ChoiYounYang2001}, it can be used as a technique for metamodelling and used in multidisciplinary optimization (MDO) frameworks. The MLSM is a weighted least squares method where the weights depend on the Euclidean distance from a sample point to where the surrogate model is to be evaluated. The weight value for a certain sample point decays as the distance increases. Describing the weight decay with a Gaussian function tends to be the most useful option even though many others have been evaluated in \cite{ToropovSchrammSahaiJones2005}. As demonstrated in \cite{PolynkinToropov2010}, the cross-validated MLSM can be used both for design variable screening and for surrogate modelling. In order to create an efficient MDO framework for problems with disparate discipline attributes \cite{OllarToropovJones2014} extended the optimization approach of MAM to the use of local DOEs and MLS approximations built in different subspaces of the total design variable space corresponding to the individual disciplines. The subspaces are finally combined into the total design variable space in which the resulting MDO problem is solved.

This paper presents a Parallel Multipoint Approximation Method that makes it capable of handling problems with the number of design variables in the order of thousands. The parallel variant of the algorithm (with the use of the shared memory) has been developed with the purpose of minimizing the worktime of the part of the algorithm related to constructing the approximation and solving the approximated problem and not related to computing the values of the objective function and  constraints. The processes of computing the values of the objective function and constraints (with the use of the 
distributed memory) are supposed to be parallelized already.

@@

It is useful to start with a brief description of MAM. A typical formulation of a constrained optimization problem that MAM works with is as follows:
\begin{equation}
  \label{eq:problem}
  \begin{array}{c}
  \min\limits_{a_i \le x_i \le b_i}F_0(x) \\
  s.t.\; F_j(x) \le 1,\; j=1,\dots ,M,
  \end{array}
\end{equation}
where $x$ is the vector of design variables, $a$ and $b$ are the lower and upper bounds for the design variables, respectively, $F_0(x)$ is the objective function, and $F_j(x)$ are the constraints. The numbers of design variables and constraints are $n$ and $M$, respectively. MAM attempts to solve this problem by using approximations of the objective function and constraints in a series of trust regions. The trust region strategy seeks to zoom in on the region where the constrained minimum is achieved. It aims at finding a trust region that is sufficiently small for the approximations to be of sufficiently good quality to improve the design, and that contains the point of the constrained minimum as its interior point. The main loop of the MAM is organized as follows.
\\

\textbf{Algorithm (MAM).}
\begin{enumerate}
\item Initialization: choose a starting point $ x^0$ and initial trust region $[ a^0,  b^0]$ such that $ x^0 \in [ a^0,  b^0]$.
\item At the $k$-th iteration the current approximation to the constrained minimum is $ x^k$, the current trust region is $[ a^k,  b^k] \subset [ a^0,  b^0]$.
  \begin{enumerate}[label=(\alph*)]
    \item Design of Experiments (DoE): a set of points $ x_k^i \in [ a^k,  b^k]$ is chosen to be used for approximation building. Responses are evaluated at the DoE points and approximations are built using the obtained values. Currently, the pool of approximation methods available in MAM consists of a metamodel assemblies \cite{PolynkinToropov2012} and the moving least-squares metamodels \cite{LancasterSalkauskas1981,Liszka1984,ChoiYounYang2001,ToropovSchrammSahaiJones2005}. Other metamodel types could be used as well.

    Denote the approximate objective function and constraints by $\widetilde{F}^k_0(x)$ and $\widetilde{F}^k_j(x)$, respectively.
    \item The original optimization problem (\ref{eq:problem}) is replaced by the following problem:
    \begin{equation}
      \label{eq:problem_approx}
      \begin{array}{c}
      \min\limits_{a_i^k \le x_i ^k\le b_i}\widetilde{F}^k_0(x) \\
      s.t.\; \widetilde{F}^k_j( x) \le 1,\; j=1,\dots ,M.
      \end{array}
    \end{equation}
    The approximate problem (\ref{eq:problem_approx}) is solved using Sequential Quadratic Programming (SQP) and the solution of this problem determines the center of the next trust region.
    \item The size of the next trust region is determined depending on the quality of approximations at the previous iteration, on the history of the points $ x^k$, and on the size of the current trust region \cite{KeulenToropovMarkine1996}.
    \item The termination criterion is checked (it is a part of the trust region strategy and depends on the position of the point $ x^{k+1}$ in the current trust region, the size of the current trust region and the quality of approximations). If the termination criterion is satisfied, the algorithm proceeds to step 3. Otherwise, it returns to step 2.
  \end{enumerate}
  \item Optimization terminates. The obtained approximation to the solution of the problem (\ref{eq:problem}) is $ x^{k+1}$.
\end{enumerate}

The selection of the approximations $\widetilde{F}^k_j(x),\; j=0, \dots ,M,$ is such that their evaluation is inexpensive as compared to the evaluation of the original response functions $F_j(x)$. For example, intrinsically linear functions were successfully used for a variety of design optimization problems in the works \cite{ToropovFilatov1993,KeulenToropov1997}. The approximations are determined by means of the weighted least squares:
\begin{equation}\label{eq:least_sq}
\min \sum_{p=1}^P{w_{pj}\left[ F_j(x_p)- \widetilde{F}^k_j(x_p,a_j) \right]^2}.
\end{equation}
In (\ref{eq:least_sq}) minimization is carried out with respect to the tuning parameters $a_j$; $w_{pj}$ are the weight coefficients, and $P$ is the number of sampling points in Design of Experiments (DoE) which must not be less than the number of parameters in the vector $a_j$.

The weight coefficients $w_{pj}$ strongly influence the difference in the quality of the approximations in different regions of the design variable space. Since in realistic constrained optimization problems the optimum point usually belongs to the boundary of the feasible region, the approximation functions should be more accurate in such domain. Thus, the information at the points located near the boundary of the feasible region is to be treated with greater weights. In a similar manner a larger weight can be allocated to a design with a better objective function, see \cite{ToropovFilatov1993,KeulenToropov1997}.

As optimization steps are carried out, a database with response function values becomes available. In order to achieve good quality approximations in the current trust region, an appropriate selection of DoE points must be made. In this work, DoE points in each trust region is generated randomly. Generally, points located far from the current trust region would not contribute to the improvement of the quality of the resulting approximations in the trust region. For this reason only points located in the neighborhood of the current trust region are taken into account, as depicted in Fig.~\ref{fig:trustregion}.
\begin{figure}[ht]
    \centering
    \includegraphics[width=0.5\textwidth]{trustregion.png}
    \caption{Current trust region (smaller box) and its extension (larger box): points outside the larger box are not used for building the approximate functions}
    \label{fig:trustregion}
\end{figure}
A box in the space of design variables, which is approximately 1.5 to 1.8 times larger than the box representing the current trust region, was found by numerical experimentation to be a reasonable choice for the size of the neighborhood.

In this work an approach is used that is based on the assembly of different approximate models $\{\varphi_l\}$ into one metamodel using the following form (note that the indices $j$ and $k$ are suppressed to simplify notation):
\begin{equation}\label{eq:assembly}
\widetilde{F}(x) = \sum_{l=1}^{NF}{b_l\varphi_l(x)},
\end{equation}
where $NF$ is the number of regressors in the model pool $\{\varphi_l\}$ and $b_l$ are corresponding regression coefficients.
The used procedure consists of two subsequent steps. In the first step, the parameters $a_l$ of individual functions (regressors) $\varphi_l$ in (\ref{eq:assembly}) are determined by solving a weighted least squares problem using a specified a DoE of $P$ points:
%\begin{equation}\label{eq:least_sq}
\[
\min \sum_{p=1}^P{w_{p}\left[ F(x_p)- \varphi_l(x_p,a_l) \right]^2},
\]
%\end{equation}
where minimization is carried out with respect to the tuning parameters $a_l$.

In the second step, based on the same DoE and keeping the obtained parameters $a_l$ fixed, a vector $b$ in (\ref{eq:assembly}) is estimated using the following formulation
\[
\min \sum_{p=1}^P{w_{p}\left[ F(x_p)- \widetilde{F}(x_p,b) \right]^2},
\]
that leads to solving a linear system of $NF$ equations with $NF$ unknowns $b_l$, where $NF$ is the number of regressors in the model pool $\{\varphi_l\}$.

The selection of the regressors $\{\varphi_l\}$ is based on the number of sampling points currently located in the trust region. In the mid-range approximation framework, inexpensive approximate models for objective and constraint functions are built using minimum required number of sampling points. The simplest case is a linear function of the tuning parameters $a$:
\[
\varphi(x) = a_0 + \sum_{i=1}^N{a_i x_i}.
\]

This structure can be extended to an \textit{intrinsically linear} function. Such functions are nonlinear, but they can be led to linear ones by simple transformations. The most useful function among them is the multiplicative function
\[
\varphi(x) = a_0 \prod_{i=1}^N{x{_i}^{a_i}}.
\]

Intrinsically linear functions have been successfully used for a variety of design optimization problems. 
The advantage of these approximation functions is that a relatively small number $N+1$ ($N$ is number of design variables) of tuning parameters $a_i$ is to be determined, and the corresponding least squares problem is solved easily. This is the most important feature of such approximations as it allows applying them to large scale optimization problems.

Other intrinsically linear functions may be considered in the model pool, e.g.
\[
\varphi(x) = a_0 + \sum_{i=1}^N{a_i/x_i},
\]
\[
\varphi(x) = a_0 + \sum_{i=1}^N{a_i x_i^2},
\]
\[
\varphi(x) = a_0 + \sum_{i=1}^N{a_i/x_i^2},
\]
\[
\varphi(x) = a_0 + \sum_{i=1}^N{a_i x_i^3},
\]
\[
\varphi(x) = a_0 + \sum_{i=1}^N{a_i/x_i^3}.
\]

As more points are added to the database the approximations may be switched to higher quality models, e.g. a rational model
\begin{equation}\label{eq:rational}
\varphi(x) = \frac{a_1 + a_2x_1 + a_3x_2+...+a_{n+1}x_n}{1 + a_{n+2}x_1 + a_{n+3}x_2+...+a_{2n+1}x_n},
\end{equation}
The coefficients in (\ref{eq:rational}) are determined using a least squares approach which reduces to a nonlinear optimization problem with a constraint on the sign of the denominator (positive or negative). The latter is necessary in order to prevent the denominator from crossing the zero axis within a specified trust region. One may note that this formulation may yield the objective function with many local minima. Currently this problem is resolved using optimization restarts from a specified number of initial guesses randomly generated in a trust region.

Tests results demonstrated that, although the above functions may describe the global behavior rather poorly, such approximations proved to be efficient in the mid-range approximation framework of MAM.

@@

Let us consider possible methods of parallelization, which could be applied for the considered problems.

First, one can parallelize the computing of the functions describing the optimized object. This way is obvious as well as necessary one since in the industrial design optimization problems the computing of even a single function value might take several hours. However, this method is a specific one for each particular problem. Here the issues of computations are addressed at the level of the application software, in which the industrial modeling is performed (e.g. Ansys, OpenFOAM, etc.).

Second, one can correct the algorithm with the purpose of parallel computing several values of the objective function and constraints at different points of the search domain. According to the MAM rules, in design of experiments $P$ 
sampling points are formed in current trust region at every iteration. The function values at these points can be computed on different processors (nodes) in parallel. The above corresponds to the parallelization of Step 2a of the algorithm with the use of the distributed memory. The number of sampling points generated within each iteration may be set equal to $P = k\cdot NP$ number, where $NP$ is a number of available processors (or nodes) and $k \geq 1$. In  terms of the time, the latter will be equivalent to $NP$ function evaluations per step. This method has been implemented successfully in \cite{PolynkinToropov2008} and has demonstrated a good efficiency since at the number of the design variables of the order of 100 major time is consumed by the function evaluations, the work of the MAM itself (in the sequential regime) introduces minor overhead.

However, when the number of variables becomes of the order of 1000, the work of the sequential part of MAM begins to affect the total problem solving time essentially (assuming the time of computing the objective function 
values remains constant). Thus, for the problem considered in Section 4 the time of execution of a single iteration of the method increased more than 4000 times (from 1.5 up to 6000 seconds) when increasing the number of variables from 100 up to  1000. Therefore, another approach to the parallelization of the algorithm has been applied within the framework of the 
present study: namely, the computational rules of MAM  providing the construction of the approximating, the solving of the approximated problem, and the choice of the next trust region (Steps 2b, 2c, and 2d of the algorithm) have been parallelized.

In order to find the most time-consuming parts of the sequential program developed earlier, we have applied Intel VTune Amplifier XE. The performed analysis has shown the matrix multiplication and solving SLAE when constructing the approximations as well as solving the approximating problem by SQP method to be the most time-consuming operations in the execution of MAM. The matrix multiplication and solving SLAE are standard operations implemented in many high-performance libraries. Here we used the corresponding parallel methods from Intel MKL library. We have parallelized the SQP method ourselves using OpenMP.

@@

The example considered in this study is a classical engineering optimization problem known as the scalable cantilevered beam \cite{Vanderpllaats2001}. The engineering object to be optimized is shown in Fig. \ref{fig:beam} taken from \cite{Vanderpllaats2001}.

@@

The design variables are the widths $b_i$ and heights $h_i$ of the segments. The number of segments $N$ can be chosen arbitrary. The total length of the beam is 500 cm, the lengths of the segments are $l_i=500/N$  cm. There are $N$ geometric constraints (the aspect ratios of each block (i.e. heights divided by widths) should not exceed 20) and $N$ constraints on the stress, calculated at the left end of each segment (stresses should not exceed  $\bar{\sigma}=14000\; N/cm^2$). There is also also a constraint on the displacement at the tip, which should not exceed 2.5 cm. The load is $P = 50 000\; N$, the Young's modulus is $E=2\cdot 10^7  N/cm^2$.

The deflection $y_i$ at the right hand of the $i$-th segment is given by the following recursive formulas:
\begin{displaymath}
  \begin{array}{c}
    y_0=y_0'=0, \\
    y_i'=\frac{P\cdot l_i}{E\cdot I_i}\Big[ L+\frac{l_i}{2}-\sum\limits_{j=1}^i l_j\Big]+y_{i-1}', \\
    y_i=\frac{P\cdot l_i^2}{2E\cdot I_i}\Big[L-\sum\limits_{j=1}^i l_j + \frac{2l_i}{3}\Big]+y_{i-1}'l_i+y_{i-1}.
  \end{array}
\end{displaymath}
The moment of inertia of a segment is $I_i={b_i h_i^3}/{12}$ and the bending moment at the left end is $M_i=P[L+l_i- \sum_{j=1}^{i}  l_j ]$. The maximum bending stress in the segment $i$ is then given by the following formula:
\begin{displaymath}
  \sigma_i=\frac{M_i h_i}{2I_i}
\end{displaymath}

>>>>

% first the title is needed
\title{Parallel Multipoint Approximation Method for Large-Scale Optimization Problems\thanks{This
study was supported by the Russian Science Foundation, project
No.\,16-11-10150.}}

@@

\author{Victor P. Gergel%(ORCID 0000-0002-4013-2329)
$^1$ \and
Konstantin A. Barkalov%(ORCID 0000-0001-5273-2471)
$^1$ \and
Evgeny A. Kozinov%(ORCID 0000-0001-6776-0096)
$^1$ \and \\
Vassili V. Toropov%(ORCID 0000-0002-7017-5055)
$^{1,2}$
}
%
\authorrunning{V.P. Gergel \and K.A. Barkalov \and E.A.Kozinov \and V.V. Toropov}
% (feature abused for this document to repeat the title also on left hand pages)

@@

\begin{abstract}
The paper presents a new development in the Multipoint Approximation Method
(MAM) that makes it capable of handling large-scale problems. The approach
relies on approximations built in the space of design variables within the
iterative trust-region-based framework of MAM. With the purpose of solving high
dimensionality problems in a reasonable time, a parallel variant of the
Multipoint Approximation Method (PMAM) has been developed. It is supposed that
the values of the objective function and those of the constraints are computed
using distributed memory (on several cluster nodes), whereas the optimization
module runs on a single node using shared memory. Numerical experiments have
been carried out on a benchmark example of structural optimization.

@@

In the present paper, the multipoint approximation method (MAM)
\cite{Toropov1989,Toropov1992,ToropovFilatov1993} and its application to
large-scale optimization problems are considered. In problems with a large (in
the order of hundreds) number of design variables, MAM has proved to be
efficient, e.g., in turbomachinery applications
\cite{ShahparPolynkinToropov2008,PolynkinToropovShahpar2008,PolynkinToropovShahpar2010}.
This method is an iterative optimization technique based on mid-range
approximations built in trust regions. A trust region is a subdomain of the
design space in which a set of design points, produced according to a
small-scale design of experiments (DoE), is evaluated. These and a subset of
previously evaluated design points are used to build metamodels of the
objective and constraint functions that are considered to be valid within a
current trust region. The trust region will then translate and change size as
optimization progresses. The trust region strategy has gone through several
stages of development to account for the presence of numerical noise in the
response function values
\cite{KeulenToropovMarkine1996,ToropovKeulenMarkine1996} and occasional
simulation failures \cite{ToropovMarkineHolden1999}. The mid-range
approximations used in the trust regions, as originally suggested in
\cite{Toropov1989} for structural optimization problems, are intrinsically
linear functions (i.e. nonlinear functions that can be reduced to a linear form
by a simple transformation) for individual substructures, and an assembly of them
for the whole structure. This was enhanced by the use of gradient-assisted
metamodels \cite{ToropovFilatov1993}, the use of simplified numerical models
which is also termed the multi-fidelity approach \cite{ToropovMarkine1996}, and
the use of analytical models derived by genetic programming
\cite{ToropovAlvarez1998}. One of the recent developments
\cite{PolynkinToropov2012} involves the use of approximation assemblies, i.e. a
two stage approximation building process that is conceptually similar to the
original one used in \cite{Toropov1989} but is free from the limitation that
lower level approximations are linked to individual substructures.

The Moving Least-Squares Method (MLSM) was proposed in
\cite{LancasterSalkauskas1981} for smoothing and interpolation of scattered
data and was later used in the mesh-free form of the finite element method
(FEM) \cite{Liszka1984}. As suggested in \cite{ChoiYounYang2001}, it can be
used as a technique for metamodeling and in multidisciplinary optimization
(MDO) frameworks. The MLSM is a weighted least-squares method where the weights
depend on the Euclidean distance from a sample point to where the surrogate
model is to be evaluated. The weight value for a certain sample point decays as
the distance increases. Describing the weight decay with a Gaussian function
tends to be the most useful option, even though many others have been evaluated
in \cite{ToropovSchrammSahaiJones2005}. As demonstrated in
\cite{PolynkinToropov2010}, the cross-validated MLSM can be used both for
design variable screening and for surrogate modeling. In order to create an
efficient MDO framework for problems with disparate discipline attributes, the
optimization approach of MAM was extended  in \cite{OllarToropovJones2014} to
the use of local DOEs and MLS approximations built in different subspaces of
the total design variable space corresponding to the individual disciplines.
The subspaces are finally combined into the total design variable space in
which the resulting MDO problem is solved.

This paper presents a Parallel Multipoint Approximation Method that makes it
capable of handling problems with numbers of design variables in the order of
thousands. The parallel variant of the algorithm (with the use of shared
memory) has been developed with the purpose of minimizing the work time of the
part of the algorithm related to constructing the approximation and solving the
approximated problem, but not related to computing the values of the objective
function and constraints. The processes of computing the values of the
objective function and constraints (with the use of distributed memory) are
supposed to be already parallelized.

@@

It would be useful to start with a brief description of MAM. A typical
formulation of a constrained optimization problem that MAM works with is as
follows:
\begin{equation}
  \label{eq:problem}
  \begin{array}{c}
  \min\limits_{a_i \le x_i \le b_i}F_0(x) \\
  \text{s.t.}\ F_j(x) \le 1,\ j=1,\dots ,M,
  \end{array}
\end{equation}
where $x$ is a vector of design variables, $a$ and $b$ are the lower and upper
bounds for the design variables, respectively, $F_0(x)$ is the objective
function, and $F_j(x)$ are the constraints. The numbers of design variables and
constraints are $n$ and $M$, respectively. MAM attempts to solve this problem
by using approximations of the objective function and constraints in a series
of trust regions. The trust region strategy seeks to zoom in on the region
where the constrained minimum is achieved. It aims at finding a trust region
that is sufficiently small for the approximations to be of sufficiently good
quality to improve the design and contains the point of the constrained minimum
as an interior point. The main loop of the MAM is organized as follows.

\subsection*{Algorithm (MAM).}
\begin{enumerate}
\item Initialization: choose a starting point $ x^0$ and initial trust
    region $[ a^0,  b^0]$ such that $ x^0 \in [ a^0,  b^0]$.

\item On the $k$th iteration, the current approximation to the constrained
    minimum is $ x^k$, and the current trust region is $[ a^k,  b^k]
    \subset [ a^0,  b^0]$.
  \begin{enumerate}[label=(\alph*)]
    \item Design of Experiments (DoE). A set of points $ x_k^i \in [
        a^k,  b^k]$ is chosen to be used for building approximations.
        Responses are evaluated at the DoE points and approximations
        are built using the obtained values. Currently, the pool of
        approximation methods available in MAM consists of metamodel
        assemblies \cite{PolynkinToropov2012} and the moving
        least-squares metamodels
        \cite{LancasterSalkauskas1981,Liszka1984,ChoiYounYang2001,ToropovSchrammSahaiJones2005}.
        Other metamodel types could be used as well.

    Denote the approximate objective function and constraints by $\widetilde{F}^k_0(x)$ and $\widetilde{F}^k_j(x)$, respectively.

    \item The original optimization problem (\ref{eq:problem}) is
        replaced by the following:
    \begin{equation}
      \label{eq:problem_approx}
      \begin{array}{c}
      \min\limits_{a_i^k \le x_i ^k\le b_i}\widetilde{F}^k_0(x) \\
      \text{s.t.}\ \widetilde{F}^k_j( x) \le 1,\ j=1,\dots ,M.
      \end{array}
    \end{equation}
    The approximate problem (\ref{eq:problem_approx}) is solved using
    Sequential Quadratic Programming (SQP). The solution of this
    problem determines the center of the next trust region.

    \item The size of the next trust region is determined depending on
        the quality of approximations on the previous iteration, on the
        history of points $ x^k$, and on the size of the current trust
        region \cite{KeulenToropovMarkine1996}.

    \item The termination criterion is checked (it is a part of the
        trust region strategy and depends on the position of the point
        $ x^{k+1}$ in the current trust region, the size of the current
        trust region and the quality of approximations). If the
        termination criterion is satisfied, the algorithm proceeds to
        step 3. Otherwise, it returns to step 2.
  \end{enumerate}
  \item Optimization terminates. The obtained approximation to the solution
      of problem (\ref{eq:problem}) is $ x^{k+1}$.
\end{enumerate}

The approximations $\widetilde{F}^k_j(x),\; j=0, \dots ,M,$ are selected in
such a way that their evaluation is inexpensive as compared to the evaluation
of the original response functions $F_j(x)$. For example, intrinsically linear
functions were successfully used for a variety of design optimization problems
in \cite{ToropovFilatov1993,KeulenToropov1997}. The approximations are
determined by means of the weighted least squares:
\begin{equation}\label{eq:least_sq}
\min \sum_{p=1}^P{w_{pj}\left[ F_j(x_p)- \widetilde{F}^k_j(x_p,a_j) \right]^2}.
\end{equation}
In (\ref{eq:least_sq}), minimization is carried out with respect to the tuning
parameters $a_j$; $w_{pj}$ are the weight coefficients, and $P$ is the number
of sampling points in Design of Experiments (DoE), which must not be less than
the number of parameters in the vector $a_j$.

The weight coefficients $w_{pj}$ strongly influence the difference in the
quality of the approximations in different regions of the design variable
space. Since in realistic constrained optimization problems the optimum point
usually belongs to the boundary of the feasible region, the approximation
functions should be more accurate in such domain. Thus, the information at the
points located near the boundary of the feasible region is to be treated with
greater weights. In a similar manner, a larger weight can be allocated to a
design with a better objective function (see
\cite{ToropovFilatov1993,KeulenToropov1997}).

As optimization steps are carried out, a database with response function values
becomes available. In order to achieve good quality approximations in the
current trust region, an appropriate selection of DoE points must be made. In
this work, DoE points in each trust region are generated randomly. Generally,
points located far from the current trust region would not contribute to the
improvement of the quality of the resulting approximations in the trust region.
For this reason, only points located in a neighborhood of the current trust
region are taken into account, as depicted in Fig.\,\ref{fig:trustregion}.
\begin{figure}[ht]
    \centering
    \includegraphics[width=0.5\textwidth]{trustregion.png}
    \caption{Current trust region (smaller box) and its extension (larger box):
    points outside the larger box are not used for building the approximate
    functions}
    \label{fig:trustregion}
\end{figure}
A box in the space of design variables, which is approximately 1.5 to 1.8 times
larger than the box representing the current trust region, was found by
numerical experimentation to be a reasonable choice for the size of the
neighborhood.

In this work, an approach is used that is based on the assembly of different
approximate models $\{\varphi_l\}$ into one metamodel using the following form
(note that the indices $j$ and $k$ are suppressed to simplify notation):
\begin{equation}\label{eq:assembly}
\widetilde{F}(x) = \sum_{l=1}^{NF}{b_l\varphi_l(x)},
\end{equation}
where $NF$ is the number of regressors in the model pool $\{\varphi_l\}$, and
$b_l$ are the corresponding regression coefficients. The  procedure used
consists of two subsequent steps. In the first step, the parameters $a_l$ of
individual functions (regressors) $\varphi_l$ in (\ref{eq:assembly}) are
determined by solving a weighted least-squares problem using a specified DoE of
$P$ points:
%\begin{equation}\label{eq:least_sq}
\[
\min \sum_{p=1}^P{w_{p}\left[ F(x_p)- \varphi_l(x_p,a_l) \right]^2},
\]
%\end{equation}
where minimization is carried out with respect to the tuning parameters $a_l$.

In the second step, based on the same DoE and keeping the obtained parameters
$a_l$ fixed, a vector $b$ in (\ref{eq:assembly}) is estimated using the
following formulation:
\[
\min \sum_{p=1}^P{w_{p}\left[ F(x_p)- \widetilde{F}(x_p,b) \right]^2},
\]
which leads to solving a linear system of $NF$ equations with $NF$ unknowns
$b_l$, where $NF$ is the number of regressors in the model pool
$\{\varphi_l\}$.

The selection of the regressors $\{\varphi_l\}$ is based on the number of
sampling points currently located in the trust region. In the mid-range
approximation framework, inexpensive approximate models for objective and
constraint functions are built using the minimum required number of sampling
points. The simplest case is that of a linear function of the tuning parameters
$a$:
\[
\varphi(x) = a_0 + \sum_{i=1}^N{a_i x_i}.
\]

This structure can be extended to an \textit{intrinsically linear} function.
Such functions are nonlinear but they can be reduced to linear ones by simple
transformations. The most useful function among them is the multiplicative
function
\[
\varphi(x) = a_0 \prod_{i=1}^N{x{_i}^{a_i}}.
\]

Intrinsically linear functions have been successfully used for a variety of
design optimization problems. The advantage of these approximation functions is
that a relatively small number $N+1$ ($N$ is the number of design variables) of
tuning parameters $a_i$ is to be determined, and the corresponding
least-squares problem is solved easily. This is the most important feature of
such approximations as it allows applying them to large-scale optimization
problems.

Other intrinsically linear functions may be considered in the model pool, e.g.,
\[
\varphi(x) = a_0 + \sum_{i=1}^N{a_i/x_i},
\]
\[
\varphi(x) = a_0 + \sum_{i=1}^N{a_i x_i^2},
\]
\[
\varphi(x) = a_0 + \sum_{i=1}^N{a_i/x_i^2},
\]
\[
\varphi(x) = a_0 + \sum_{i=1}^N{a_i x_i^3},
\]
\[
\varphi(x) = a_0 + \sum_{i=1}^N{a_i/x_i^3}.
\]

As more points are added to the database, the approximations may be switched to
higher quality models, e.g., a rational model
\begin{equation}\label{eq:rational}
\varphi(x) = \frac{a_1 + a_2x_1 + a_3x_2+...+a_{n+1}x_n}{1 + a_{n+2}x_1 + a_{n+3}x_2+...+a_{2n+1}x_n}.
\end{equation}
The coefficients in (\ref{eq:rational}) are determined using a least-squares
approach which reduces to a nonlinear optimization problem with a constraint on
the sign of the denominator (positive or negative). The latter is necessary in
order to prevent the denominator from crossing the zero axis within a specified
trust region. One may note that this formulation may yield an objective
function with many local minima. Currently, this problem is resolved using
optimization restarts from a specified number of initial guesses randomly
generated in a trust region.

Tests results demonstrated that, although the above functions may describe the
global behavior rather poorly, such approximations prove to be efficient in the
mid-range approximation framework of MAM.

@@

Let us consider possible methods of parallelization that could be applied to
the problems considered.

First, one can parallelize the computation of the functions describing the
optimized object. This way is an obvious as well as necessary one since, in
industrial design optimization problems, the computation of even a single
function value may take several hours. However, this method is a specific one
for each particular problem. Here the computation issues are addressed at the
level of the application software in which the industrial modeling is performed
(e.g., Ansys, OpenFOAM, etc.).

Second, one can correct the algorithm with the purpose of parallelly computing
several values of the objective function and constraints at different points of
the search domain. According to the MAM rules, in design of experiments, $P$
sampling points are formed in the current trust region at each iteration. The
function values at these points can be computed on different processors (nodes)
in parallel. The above corresponds to the parallelization of Step 2a of the
algorithm using distributed memory. The number of sampling points generated
within each iteration may be set equal to $P = k\cdot NP$, where $NP$ is the
number of available processors (or nodes), and $k \geq 1$. In terms of time,
the latter will be equivalent to $NP$ function evaluations per step. This
method has been implemented successfully in \cite{PolynkinToropov2008} and has
demonstrated a good efficiency since, for numbers of design variables of the
order of 100, time is mainly consumed by the function evaluations, whereas the
work of the MAM itself (in the sequential regime) introduces a minor overhead.

However, when the number of variables becomes of the order of 1000, the work of
the sequential part of MAM begins to affect the total problem solving time
essentially (assuming that the time of computing the objective function values
remains constant). Thus, for the problem considered in Section 4, the time of
execution of a single iteration of the method increased by a factor of more
than 4000 (from 1.5 up to 6000 seconds) when increasing the number of variables
from 100 up to 1000. Another approach to the parallelization of the algorithm
has therefore been applied within the framework of the present study: namely,
the computational rules of MAM  providing for the construction of the
approximation, the solution of the approximated problem, and the choice of the
next trust region (Steps 2b, 2c, and 2d of the algorithm) were parallelized.

In order to find the most time-consuming parts of the sequential program
developed earlier, we applied the Intel VTune Amplifier XE. The analysis
performed has shown that the most time-consuming operations in the execution of
MAM are matrix multiplication, the solution of the SLAE when constructing the
approximations, and the solution of the approximating problem by SQP. Matrix
multiplication and solution of SLAE are standard operations implemented in many
high-performance libraries. Here we used the corresponding parallel methods
from the Intel MKL library. We parallelized the SQP method ourselves using
OpenMP.

@@

The example considered in this study is a classical engineering optimization
problem known as the scalable cantilevered beam \cite{Vanderpllaats2001}. The
engineering object to be optimized is shown in Fig.\,\ref{fig:beam} (taken from
\cite{Vanderpllaats2001}).

@@

The design variables are the widths $b_i$ and heights $h_i$ of the segments.
The number of segments $N$ can be chosen arbitrarily. The total length of the
beam is 500~cm, the lengths of the segments are $l_i=500/N$~cm. There are $N$
geometric constraints (the aspect ratios of the blocks, i.e. heights divided by
widths, should not exceed 20) and $N$ constraints on the stress, calculated at
the left end of each segment (stresses should not exceed
$\bar{\sigma}=14\,000\;\text{N}/\text{cm}^2$). There is also a constraint on
the displacement at the tip, which should not exceed 2.5~cm. The load is $P =
50\,000$\;N; the Young's modulus is $E=2\cdot 10^7\;\text{N}/\text{cm}^2$.
